% -*- mode: latex; fill-column: 65; -*-
\documentclass[aps, pra, preprint]{revtex4-1}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{blindtext}
\presetkeys%
    {todonotes}%
    {inline,backgroundcolor=yellow}{}

\frenchspacing

\bibliographystyle{apsrev4-1}

\begin{document}

\title{Determination of temperatures of the axial modes of an ion
  crystal in a Penning trap}
\author{Dominic Meiser}
\affiliation{Trimble Inc, Boulder, 4730 Walnut Street, Suite 201,
  Boulder, CO 80301, USA.}

\begin{abstract}
  In these notes we discuss how to extract the temperatures of
  the drum-head modes of a crystal of ultra-cold ions in a
  Penning trap from spectra of the axial motion.
\end{abstract}

\maketitle


\section{The drum head modes of an ion crystal}

We consider an ensemble of $N$ ions in a Penning trap with the
magnetic field aligned with the $z$ axis. We assume that the
motion of the ions along the $z$ direction is decoupled from the
other degrees of freedom. We consider the $z$ degrees of
freedom only.

The Hamiltonian is given by
\begin{equation}
  H=
    \sum_{j=1}^N\frac{p_j^2}{2m_j} +
    \sum_{j=1}^NV_1(z_j) +
    \sum_{\substack{i,j = 1\\i\neq j}}^{N}V_2(z_i - z_j)\;.
\end{equation}
In this equation, $p_j$ is the $z$-component of the momentum of
ion $j$. The single particle potential $V_1$ represents the
electrostatic trap potentials, and the pairwise interactions
$V_2$ are due to the Coulomb interaction. 

Expanding the Hamiltonian to second order around the stable
steady state at $z_j=0$, we obtain
\begin{equation}
  H\approx 
    \sum_{j=1}^N \frac{p_j^2}{2m_j} +
    \frac{1}{2}\sum_{j=1}^N m_j \omega_z^2 z_j^2+
    \frac{1}{2}\sum_{\substack{i,j = 1\\i\neq j}}^{N} V_{ij}(z_i-z_j)^2\;.
\end{equation}

Introducing the vectors
\begin{equation}
  z=\left[
    \begin{array}{c}
      z_1\\
      z_2\\
      \vdots
    \end{array}
  \right]
\end{equation}
and
\begin{equation}
  p=\left[
    \begin{array}{c}
      p_1\\
      p_2\\
      \vdots
    \end{array}
  \right]
\end{equation}
along with the matrices
\begin{equation}
  M=\left[
    \begin{array}{ccc}
      m_1&0&\hdots\\
      0&m_2&\\
      \vdots&&\ddots
    \end{array}
  \right]
\end{equation}
and
\begin{equation}
  V=\left[
    \begin{array}{cccc}
      m_1\omega_z^2&V_{1,2}&V_{1,3}&\hdots\\
      V_{2,1}&m_2\omega_z^2&V_{2,3}&\hdots\\
      V_{3,1}&V_{3,2}&m_3\omega_z^2&\\
      \vdots&\vdots&&\ddots
    \end{array}
  \right]\;,
\end{equation}
we can write the linearized Hamiltonian as
\begin{equation}
  H =
  \frac{1}{2}\left\langle p, M^{-1}p \right\rangle +
  \frac{1}{2}\left\langle z, V z \right \rangle\;,
  \label{eqn:linearized_hamiltonian}
\end{equation}
with $\langle \cdot, \cdot\rangle$ the inner product.

The equations of motion are
\begin{eqnarray}
  \dot{z} &=& \frac{\partial H}{\partial p} = M^{-1}p\label{eqn:zdot}\\
  \dot{p} &=& -\frac{\partial H}{\partial z} = -Vz\label{eqn:pdot}\;.
\end{eqnarray}
Taking another time derivative of Eqn.~\eqref{eqn:zdot} and
eliminating $\dot{p}$ using Eqn.~\eqref{eqn:pdot} we find
\begin{equation}
  M\ddot{z}=-Vz\;.
\end{equation}
Making the ansatz
\begin{equation}
  z(t)=e^{i\omega t}z(0)
\end{equation}
we obtain the eigenvalue equation
\begin{equation}
  M\omega^2z = Vz\;.
\end{equation}
Since $M$ and $V$ are symmetrical we have a complete orthogonal
set of eigenvectors $\phi_1, \phi_2,\ldots$ with eignvalues
$\omega_1, \omega_2, \ldots$. We assume that the eigenmodes are
normalized according to
\begin{equation}
  \left\langle \phi_m,\phi_n\right \rangle = a^2\delta_{m,n}\;,
\end{equation}
where $a$ is the unit of length and $\delta_{m,n}$ is Kronecker's
delta symbol.


\section{Energy in terms of eigenmodes}

We can insert the expansion of the ion trajectories in terms of the
eigenmodes,
\begin{equation}
  z=\sum_{m=1}^N e^{i\omega_m t} \alpha_m \phi_m\;,
\end{equation}
into the Hamiltonian Eqn.~\eqref{eqn:linearized_hamiltonian}. A
simple calculation yields
\begin{equation}
  H=\sum_{m,n=1}^Na^2\left| \alpha_m\right|^2
    \omega_m^2\left\langle \phi_m,M\phi_n \right\rangle\;.
\end{equation}
This can be further simplified in the case where all ions
have the same mass, $m=m_1=m_2=\ldots$. We now specialize our
discussion to that case. We obtain
\begin{equation}
  H=ma^2\sum_{m=1}^N\omega_m^2\left| \alpha_m \right|^2\;.
\end{equation}
Each mode contributes
\begin{equation}
  E_m=ma^2\omega_m^2\left| \alpha_m \right|^2\;,
\end{equation}
to the total energy. Note that if we were to choose the harmonic
oscillator length, $a_{\rm osc}=\sqrt{\hbar/(m\omega_m)}$ as the
unit of length this reduces to the well known formula
\begin{equation}
  E_m=\hbar\omega_m\left| \alpha_m \right|^2\;.
\end{equation}
But the choice of unit of length is immaterial for our discussion
here.


\section{Determination of mode energies from spectra}

From the dynamical simulations we compute the power spectral
density
\begin{equation}
  I(\omega) =
    \sum_{j=1}^N\left| \tilde{z}_j(\omega) \right|^2 +
    \sum_{j=1}^N\left| \tilde{z}_j(-\omega) \right|^2 =
    \left\langle \tilde{z}(\omega), \tilde{z}(\omega)\right \rangle +
    \left\langle \tilde{z}(-\omega), \tilde{z}(-\omega)\right \rangle 
    \;.\label{eqn:psd}
\end{equation}
In this equation, $\tilde{z}_j$ is the Fourier transform of the
trajectory of ion $j$,
\begin{equation}
  \tilde{z}_j(\omega) = \frac{1}{T}\int_0^T e^{-i\omega t}z_j(t)\, dt,
\end{equation}
with $T$ the total integration time of the simulation. Inserting
our mode expansion the power spectral density becomes
\begin{equation}
  I(\omega) =
  \sum_{m=1}^N\frac{a^2 \left| \alpha_m \right|^2}{T^2}
  \int_0^T\int_0^T\left( e^{-i(\omega-\omega_m)(t-t^\prime)} +
    e^{i(\omega+\omega_m)(t-t^\prime} \right)\,dt\,dt^\prime\;.
\end{equation}
We now consider a single resonance $\omega_m$ and assume that
its resonance is well resolved. In that case we can integrate the
power spectral density over the resonance. We find
\begin{equation}
  \int_{\omega_m-\Delta}^{\omega_m+\Delta} I(\omega)\,d\omega =
  \frac{\pi a^2}{T}\left| \alpha_m \right|^2\;,
\end{equation}
where we have assume that $\Delta\gg T^{-1}$. Solving for
$|\alpha_m|^2$ and inserting in the expression for the mode
energy we find
\begin{equation}
  E_m = m\omega_m^2\frac{T}{\pi}
        \int_{\omega_m-\Delta}^{\omega_m+\Delta}I(\omega)\,d\omega\;.
\end{equation} 


\section{Note on discrete signals and discrete Fourier transform}

In the simulations we sample the ion motion with a discrete,
constant sampling rate $\Delta t$. In terms of this discretely
sampled signal, the Fourier transform can be approximate by a
discrete Fourier transform (which we evaluate with the aid of a
Fast Fourier Transform). We can consider the discrete Fourier
transform as a Riemann sum approximation to the Fourier
transform,
\begin{equation}
  \tilde{z}(\omega)=
  T^{-1}\int_0^Te^{-i\omega t}z(t)\,dt \approx
  T^{-1}\sum_n^{N_s}\Delta t e^{-i\omega n\Delta t}z(n \Delta t) =
  \frac{1}{N_s}\sum_n^{N_s}e^{-i\omega n\Delta t}z(n \Delta t)\;.
\end{equation}
In this equation, $N_s = T / \Delta t$ is the total number of
samples. Introducing the frequency grid spacing
$\Delta\omega=2\pi/T$ we can write the approximate Fourier
transform as
\begin{equation}
  \tilde{z}(l\Delta\omega)\approx
  \frac{1}{N_s}\sum_n^{N_s}e^{-i2\pi ln/N_s}z(n \Delta t)\;.
\end{equation}
We have to keep in mind that frequencies exceeding the Nyquist
frequency, $l\Delta\omega>\pi/\Delta t$, wrap around to negative
frequencies.


\end{document}
